sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-02 doc/paper.Rmd
Edit intro

---
title: "`sdmTMB`: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], Other authors\n| $^1$Pacific Biological Station Fisheries and Oceans Canada, \n| Nanaimo, BC, Canada\n|
  $^2$Other addresses\n"
output:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent Point 1: set the context for and purpose of the work; \smallskip 
    | Point 2: indicate the approach and methods; \smallskip 
    | Point 3: outline the main results; \smallskip 
    | Point 4: identify the conclusions and the wider implications. \smallskip 
---

@@

Ecological data are often collected in space or in space repeatedly over time. 
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model [e.g., @cressie1993; @cressie2011]. 
Data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables. 
While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation. 
This residual correlation is ideally dealt with because doing so (1) results in valid statistical inference [e.g., @legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @barnett2021].

Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data. 
Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial/spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time. 
A common approach to modelling such random effects is with Gaussian Random fields (GRFs), where the random effects are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].

While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009]. 
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016]. 
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian or maximum likelihood methods.
This can make an enormous difference to computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA.
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

@@

Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB and INLA in a fast, flexible, and user-friendly interface that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a]. One common application is for species distribution models (SDMs), hence the package name. This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through X case studies, and concludes with a discussion of additional functionality and related software.

-   [@anderson2019]
-   [@maureaud2021]
-   [@anderson2020]
-   [@osgood-zimmerman2021]
-   [@breivik2021]
-   [@monnahan2021]
-   [@shelton2014]
-   [@lindgren2011]
-   [@latimer2009]

Packages:

spaMM: @rousset2014
spNNGP: @finley2021

millard1985: Space–Time Correlation and Its Effects on Methods for Detecting Aquatic Ecological Change "In practice, spatial correlation is more likely to be a problem than is temporal correlation, given typical monitoring frequencies."

legendre1989a: "Spatial pattern and ecological analysis": "We can expect the spatial approach to ecological problems to bring about a quantum jump for ecologists and population geneticists who had learned a type of statistics where one had to hide space or time structures. It is now possible to use these structures and to integrate them into our analyses as fully-fledged controlled variables."

ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany.

>>>>

---
title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], Other authors\n| $^1$Pacific Biological Station Fisheries and Oceans Canada, \n| Nanaimo, BC, Canada\n|
  $^2$Other addresses\n"
output:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent Point 1: set the context for and purpose of the work; \smallskip 
    | Point 2: indicate the approach and methods; \smallskip 
    | Point 3: outline the main results; \smallskip 
    | Point 4: identify the conclusions and the wider implications. \smallskip 
---

@@

Ecological data are often collected in space or in space repeatedly over time. 
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [e.g., @cressie1993; @cressie2011]. 
While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation. 
This residual correlation is ideally dealt with because doing so (1) results in valid statistical inference [e.g., @legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @barnett2021].

Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data. 
Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial/spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time. 
A common approach to modelling such random effects is with Gaussian random fields (GRFs), where the random effects are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].

While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009]. 
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016]. 
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian or maximum likelihood methods.
This can make an enormous difference to computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

@@

Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB and INLA.
Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
One common application is for species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through X case studies, and concludes with a discussion of additional functionality and related software.

<!-- -   [@anderson2019] -->
<!-- -   [@maureaud2021] -->
<!-- -   [@anderson2020] -->
<!-- -   [@osgood-zimmerman2021] -->
<!-- -   [@breivik2021] -->
<!-- -   [@monnahan2021] -->
<!-- -   [@shelton2014] -->
<!-- -   [@lindgren2011] -->
<!-- -   [@latimer2009] -->

<!-- Packages: -->

<!-- spaMM: @rousset2014 -->
<!-- spNNGP: @finley2021 -->

<!-- millard1985: Space–Time Correlation and Its Effects on Methods for Detecting Aquatic Ecological Change "In practice, spatial correlation is more likely to be a problem than is temporal correlation, given typical monitoring frequencies." -->

<!-- legendre1989a: "Spatial pattern and ecological analysis": "We can expect the spatial approach to ecological problems to bring about a quantum jump for ecologists and population geneticists who had learned a type of statistics where one had to hide space or time structures. It is now possible to use these structures and to integrate them into our analyses as fully-fledged controlled variables." -->

<!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->
